---
title: "microrefugia review"
format: revealjs
editor: visual
---


```{r}
library(tidyverse)

d <- 
  readxl::read_xlsx('data/Microrefugia literature review table - data collection.xlsx', 
                    sheet = 1,
                    skip = 1) |>
  janitor::clean_names() |>
  slice(2:999) |>
  dplyr::select(-2)
```

## screening

```{r}
screening <- readxl::read_xlsx('data/Microrefugia review studies.xlsx', skip = 1) |>
  janitor::clean_names()

# included versus not
paste('included versus not:')
screening$included |> table()
```



## reasons for exclusion

```{r}
screening |>
  dplyr::filter(included == "No") |>
  dplyr::select(value = reason_for_exclusion) |>
  dplyr::mutate(value = str_replace_all(value, '\\(e.g. disturbance, nitrogen\\)', '') |>
                  str_replace_all("\\'", "") |>
                  str_replace_all('-- were only loking at climate, temperature, moisture, microclimate', '') |>
                  str_replace_all('\\(were not covering palaeo, pleistocene divergence\\)', '')) |> 
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Reason for Exclusion")
```

## included by year

```{r}
screening |>
  dplyr::filter(!is.na(reader)) |>
  tidyr::replace_na(list(included = 'Left Blank')) |>
  dplyr::group_by(included, year) |>
  dplyr::summarise(n = n()) |>
  dplyr::ungroup() |> 
  rbind(expand_grid(included = c('No', "Yes", 'Left Blank'), year = 2005:2026, n=0)) |> 
  group_by(included, year) |>
  summarise(n=sum(n)) |>
  ggplot(aes(x=year, y=n, fill = included)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  scale_fill_brewer(palette = 'Set2')

```

## Many locations, mostly defined

```{r}
# locations - could be a map (requires one person to get lat/longs for all 200+ studies)
d$study_location |> unique() |> length() |> paste('locations')

# yes or no questions
paste('is the refugia defined:')
d$is_the_specified_refugia_defined |> table()

```

## habitats


```{r}
d |>
  dplyr::select(habitat) |>
  tidyr::separate(habitat, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit() |>
  dplyr::mutate(value = str_remove_all(value, "[^A-Za-z0-9]")) |>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  arrange(desc(n)) |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value, n))) +
  geom_bar(stat = 'identity') +
  ggtitle("Habitat")
```

## species groups

```{r}
d |>
  dplyr::select(species_group) |>
  tidyr::separate(species_group, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Species Group")
```

## framework 

```{r}
d |>
  dplyr::select(framework) |>
  tidyr::separate(framework, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Framework (Some Studies cover multiple categories)")
```

## refugia term

```{r}
d |>
  dplyr::select(refugia_term) |>
  tidyr::separate(refugia_term, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  mutate(value = str_to_lower(value)) |>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Refugia Term (Some Studies cover multiple categories)")
```

## reference for definition

This needs to be cleaned up (i.e. standardized) to make it easier to summarize

```{r}
d$refugia_defintion_reference |> table() |> sort()
```

## position in relation to main range

```{r}
d |> 
  group_by(y= position_in_relation_to_main_range) |>
  summarise(n = n()) |>
  ungroup() |>
  mutate(y = ifelse(y == 'NA', "NA (entered as such)", y)) |>
  tidyr::replace_na(list(y = "NA (because it was left blank)")) |>
  ggplot(aes(x=n, y=fct_reorder(y,n))) +
  geom_bar(stat = 'identity') +
  ggtitle('Position in relation to main range')
```

## sampling structure

```{r}
d |>
  dplyr::select(sampling_structure) |>
  tidyr::separate(sampling_structure, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  # mutate(value = str_to_lower(value),
  #        value = ifelse(value == 'na', 'entered as NA', value)) |>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Sampling Structure (Some Studies cover multiple categories)")
```


## climate driver

```{r}
d |>
  dplyr::select(value = what_kind_of_climate_driver_does_the_study_want_to_capture_ie_which_aspect_is_the_refugia_buffering) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>

  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Climate Driver (Some Studies cover multiple categories)")
```

## exposure_considered

```{r}
d |>
  dplyr::select(value = exposure_considered)  |> 
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Exposure Considered (Some Studies cover multiple categories)")
```

## species_sensitivity

ie_what_is_the_vulnerable_part_of_the_species

```{r}

d |>
  dplyr::select(value = species_sensitivity_ie_what_is_the_vulnerable_part_of_the_species)  |> 
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Species Sensitivity (Some Studies cover multiple categories)")
```

## Species sensitivity used to ID refugia

```{r}

d |>
  dplyr::select(value = is_species_sensitivity_used_to_identify_refugia) |> 
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)"))  |> 
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit() |>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("is Species Sensitivity used to identigy refugia")
```

## Species adaptablity

```{r}
# d$species_adaptability_genetic_adaptability_and_dispersal |> table()
d |>
  dplyr::select(value = species_adaptability_genetic_adaptability_and_dispersal) |> 
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("species_adaptability_genetic_adaptability_and_dispersal")
```

## refugia_structure

```{r}
# d$ |> table()
d |>
  dplyr::select(value = refugia_structure_extrinsic_adaptability_factors) |> 
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("refugia_structure_extrinsic_adaptability_factors")
```

## species response to changing climate

```{r}
# d$species_response_to_changing_climate |> table()
d |>
  dplyr::select(value = species_response_to_changing_climate) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("species_response_to_changing_climate")


```


## plot size

```{r}
d |>
  dplyr::select(value = in_situ_plot_size) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("In situ plot size")
```


## spatial resolution

```{r}
d |>
  dplyr::select(value = spatial_resolution_modelling_rs) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("spatial_resolution_modelling_rs")

```

## spatial extent


```{r}
d |>
  dplyr::select(value = spatial_extent) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("spatial extent")

```

## temporal aspects

```{r}
d |>
  dplyr::select(value = temporal_aspects) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Temporal aspects")

```

## climate variables

```{r}
d |>
  dplyr::select(value = climate_variables) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Climate variables")
```


## 


```{r}
d |>
  dplyr::select(value = climate_data_source) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("Climate data source")
```

##

```{r}
d |>
  dplyr::select(value = climate_data_resolution) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("climate_data_resolution")
```

##

```{r}
d |>
  dplyr::select(value = other_habitat_variables) |>
  mutate(value = str_to_lower(value),
         value = ifelse(value == 'na', 'entered as NA', value)) |>
  tidyr::replace_na(list(value = "NA (because it was left blank)")) |>
  tidyr::separate(value, into = c('w','x', 'y', 'z', 'a', 'b'), sep = '\\, ') |>
  tidyr::pivot_longer( c('w','x', 'y', 'z', 'a', 'b'))|> 
  na.omit()|>
  group_by(value) |>
  summarise(n = n()) |>
  ungroup() |>
  dplyr::mutate(value = ifelse(n == 1, 'other', value)) |>
  ggplot(aes(x=n, y=fct_reorder(value,n))) +
  geom_bar(stat= 'identity') +
  ggtitle("other habitat variables")
```

## specific quantification



```{r}
paste(
d$specific_quantification |> is.na() |> sum(), 'descriptions entered')

```
